---
title: "scRNA-seq Cluster Proportion Modeling and Simulation-Based Power"
author: "GW McElfresh"
date: "2026-04-26"
output:
  html_document:
    toc: true
    toc_depth: 3
---

## Goals

This notebook does four things:

1. Simulates a small scRNA-seq-like dataset (3 subjects, 2 conditions each, 300 genes).
2. Loads a real Cellxgene `.h5ad` file and converts metadata into subject-by-treatment-by-cluster proportional data.
3. Fits Bayesian proportion models in R using `brms` with the `cmdstanr` backend.
4. Uses the fitted distributional parameters as anchors for simulation-based power grids with two-sided Wilcoxon signed-rank tests.

## 1) Data Source and Download

Integrated cancer cell-specific single-cell RNA-seq datasets of immune checkpoint blockade-treated patients can be downloaded from Cellxgene:

- https://datasets.cellxgene.cziscience.com/134d34af-cbcd-4837-9310-3d1f83ec6f18.h5ad

You can download the file from a terminal:

```bash
mkdir -p data
curl -L "https://datasets.cellxgene.cziscience.com/134d34af-cbcd-4837-9310-3d1f83ec6f18.h5ad" -o data/icb_integrated.h5ad
```

If needed, update the path in the Python chunk below.

## 2) Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(purrr)
library(stringr)
library(tibble)

library(brms)
library(cmdstanr)

set.seed(20260426)
options(mc.cores = max(1, parallel::detectCores() - 1))
options(brms.backend = "cmdstanr")
```

Install once (if needed):

```{r install-pkgs, eval=FALSE}
install.packages(c("dplyr", "tidyr", "ggplot2", "readr", "purrr", "stringr", "tibble", "brms", "cmdstanr"))
cmdstanr::install_cmdstan()
```

## 3) Simulate a Basic scRNA-seq-like Dataset

This creates cell-level metadata and a 300-gene expression matrix for 3 subjects, each measured in 2 conditions.

```{r simulate-scrna}
subjects <- paste0("S", 1:3)
conditions <- c("control", "treated")
clusters <- paste0("C", 1:6)
ngenes <- 300

cells_per_subject_condition <- 800
n_total <- length(subjects) * length(conditions) * cells_per_subject_condition

meta_sim <- tidyr::expand_grid(
  subject = subjects,
  condition = conditions,
  cell_index = seq_len(cells_per_subject_condition)
) %>%
  mutate(
    # Mild treatment enrichment in C3/C4
    cluster = purrr::map_chr(condition, function(cc) {
      if (cc == "treated") {
        sample(clusters, size = 1, prob = c(0.14, 0.14, 0.26, 0.22, 0.14, 0.10))
      } else {
        sample(clusters, size = 1, prob = c(0.20, 0.20, 0.18, 0.16, 0.16, 0.10))
      }
    })
  )

# Simulate a simple gene expression matrix with cluster-dependent means.
# Rows: cells, Columns: genes.
cluster_effect <- setNames(c(0.8, 0.9, 1.2, 1.1, 0.95, 0.85), clusters)
expr_sim <- matrix(rpois(n_total * ngenes, lambda = 1), nrow = n_total, ncol = ngenes)
for (k in seq_len(n_total)) {
  lam <- cluster_effect[meta_sim$cluster[k]]
  expr_sim[k, ] <- rpois(ngenes, lambda = lam)
}
colnames(expr_sim) <- paste0("G", seq_len(ngenes))

head(meta_sim)
summary(as.vector(expr_sim))
```

Convert simulated metadata to cluster proportions:

```{r simulated-proportions}
prop_sim <- meta_sim %>%
  count(subject, condition, cluster, name = "n_cells") %>%
  group_by(subject, condition) %>%
  mutate(total_cells = sum(n_cells), prop = n_cells / total_cells) %>%
  ungroup()

prop_sim
```

## 4) Load Cellxgene .h5ad and Build Proportion Data (Python)

This chunk reads the `.h5ad`, detects likely metadata columns for subject/treatment/cluster, and writes a CSV used by R.

```{python build-proportions-python}
import os
import numpy as np
import pandas as pd

h5ad_path = "data/icb_integrated.h5ad"
out_csv = "analysis/cellxgene_cluster_proportions.csv"

if not os.path.exists(h5ad_path):
    raise FileNotFoundError(
        f"Could not find {h5ad_path}. Download the file first or update h5ad_path."
    )

import anndata as ad
adata = ad.read_h5ad(h5ad_path)
obs = adata.obs.copy()
obs.columns = [str(c) for c in obs.columns]

# Heuristic selection for key columns
subject_candidates = [
    "donor_id", "patient_id", "subject_id", "sample_id", "donor", "patient", "subject"
]
treatment_candidates = [
    "treatment", "condition", "therapy", "response", "timepoint", "group"
]
cluster_candidates = [
    "cell_type", "celltype", "cell_type_ontology_term_id", "cluster", "leiden", "annotation"
]

def pick_column(columns, candidates):
    lower_map = {c.lower(): c for c in columns}
    for cand in candidates:
        if cand.lower() in lower_map:
            return lower_map[cand.lower()]
    return None

subject_col = pick_column(obs.columns, subject_candidates)
treatment_col = pick_column(obs.columns, treatment_candidates)
cluster_col = pick_column(obs.columns, cluster_candidates)

if subject_col is None or treatment_col is None or cluster_col is None:
    print("Available obs columns:")
    print(list(obs.columns))
    raise ValueError(
        "Could not auto-detect subject/treatment/cluster columns. "
        "Edit candidates or set subject_col/treatment_col/cluster_col manually."
    )

print("Using columns:")
print("subject:", subject_col)
print("treatment:", treatment_col)
print("cluster:", cluster_col)

meta = obs[[subject_col, treatment_col, cluster_col]].copy()
meta.columns = ["subject", "treatment", "cluster"]
meta = meta.dropna(subset=["subject", "treatment", "cluster"])

counts = (
    meta
    .groupby(["subject", "treatment", "cluster"], as_index=False)
    .size()
    .rename(columns={"size": "n_cells"})
)

totals = (
    counts
    .groupby(["subject", "treatment"], as_index=False)["n_cells"]
    .sum()
    .rename(columns={"n_cells": "total_cells"})
)

prop_df = counts.merge(totals, on=["subject", "treatment"], how="left")
prop_df["prop"] = prop_df["n_cells"] / prop_df["total_cells"]

os.makedirs("analysis", exist_ok=True)
prop_df.to_csv(out_csv, index=False)
print(f"Saved proportion dataframe: {out_csv}")
print(prop_df.head())
```

## 5) Read Proportion Data in R and Prepare Modeling Table

```{r read-proportion-csv}
prop_real <- readr::read_csv("analysis/cellxgene_cluster_proportions.csv", show_col_types = FALSE)

# Ensure factors and add a beta-safe transformed proportion in (0,1)
prop_real <- prop_real %>%
  mutate(
    subject = as.factor(subject),
    treatment = as.factor(treatment),
    cluster = as.factor(cluster)
  )

n_eps <- 0.5
prop_real <- prop_real %>%
  mutate(prop_beta = (n_cells + n_eps) / (total_cells + 2 * n_eps))

dplyr::glimpse(prop_real)
```

Optional quick metadata summaries:

```{r summaries}
by_subject <- prop_real %>% count(subject, name = "rows")
by_treatment <- prop_real %>% count(treatment, name = "rows")
by_cluster <- prop_real %>% count(cluster, name = "rows")

by_subject
by_treatment
by_cluster
```

## 6) Bayesian Proportion Models with brms + cmdstanr

### Model choice

- If proportions include exact 0 or 1 frequently, use `zero_one_inflated_beta()`.
- If not, standard `beta()` with continuity correction (`prop_beta`) is often stable.

Below, we model one cluster at a time and estimate treatment effect plus subject-level variance.

```{r fit-models, message=FALSE, warning=FALSE}
# Keep clusters with enough observations to avoid unstable toy fits.
candidate_clusters <- prop_real %>%
  count(cluster, name = "n") %>%
  filter(n >= 8) %>%
  pull(cluster)

# Priors are intentionally weakly informative for a basic workflow.
priors_beta <- c(
  set_prior("normal(0, 1)", class = "b"),
  set_prior("normal(0, 1)", class = "Intercept"),
  set_prior("student_t(3, 0, 1)", class = "sd"),
  set_prior("gamma(2, 0.1)", class = "phi")
)

fit_one_cluster <- function(cl) {
  dat <- prop_real %>% filter(cluster == cl)
  if (nrow(dat) < 8) return(NULL)

  brm(
    formula = bf(prop_beta ~ treatment + (1 | subject)),
    data = dat,
    family = Beta(),
    prior = priors_beta,
    chains = 4,
    iter = 2000,
    warmup = 1000,
    seed = 20260426,
    refresh = 0
  )
}

fits <- setNames(vector("list", length(candidate_clusters)), candidate_clusters)
for (cl in candidate_clusters) {
  cat("Fitting cluster", as.character(cl), "...\n")
  fits[[as.character(cl)]] <- fit_one_cluster(cl)
}

# Keep successful fits only
fits <- fits[!vapply(fits, is.null, logical(1))]
length(fits)
```

Extract treatment effect and subject-level SD:

```{r extract-params}
extract_summary <- function(fit, cluster_name) {
  fx <- posterior_summary(fit, pars = c("b_treatmenttreated", "sd_subject__Intercept", "phi"))
  tibble(
    cluster = cluster_name,
    term = rownames(fx),
    estimate = fx[, "Estimate"],
    q2.5 = fx[, "Q2.5"],
    q97.5 = fx[, "Q97.5"]
  )
}

param_summaries <- purrr::imap_dfr(fits, ~extract_summary(.x, .y))
param_summaries
```

## 7) Simulation-Based Power Grids (Two-Sided Wilcoxon)

Idea:

1. Pick a baseline mean proportion and variance scale.
2. Simulate paired subject-level control/treated proportions from beta distributions with a subject random effect.
3. Test treated vs control with two-sided paired Wilcoxon.
4. Repeat many times in each grid bin to estimate power.

```{r power-grid-functions}
inv_logit <- function(x) 1 / (1 + exp(-x))

simulate_one_cohort <- function(
  n_subject,
  mu_control,
  mu_treated,
  phi,
  sd_subject,
  seed = NULL
) {
  if (!is.null(seed)) set.seed(seed)

  # Random intercept shared across both conditions for each subject
  b_subj <- rnorm(n_subject, mean = 0, sd = sd_subject)

  # Condition means on logit scale with subject shift
  eta_c <- qlogis(mu_control) + b_subj
  eta_t <- qlogis(mu_treated) + b_subj

  mu_c <- pmin(pmax(inv_logit(eta_c), 1e-6), 1 - 1e-6)
  mu_t <- pmin(pmax(inv_logit(eta_t), 1e-6), 1 - 1e-6)

  a_c <- mu_c * phi
  b_c <- (1 - mu_c) * phi
  a_t <- mu_t * phi
  b_t <- (1 - mu_t) * phi

  y_c <- rbeta(n_subject, shape1 = a_c, shape2 = b_c)
  y_t <- rbeta(n_subject, shape1 = a_t, shape2 = b_t)

  list(control = y_c, treated = y_t)
}

power_in_bin <- function(
  n_subject,
  mu_control,
  mean_shift,
  phi,
  sd_subject,
  n_sims = 500,
  alpha = 0.05
) {
  mu_treated <- pmin(pmax(mu_control + mean_shift, 0.001), 0.999)

  pvals <- replicate(n_sims, {
    dat <- simulate_one_cohort(
      n_subject = n_subject,
      mu_control = mu_control,
      mu_treated = mu_treated,
      phi = phi,
      sd_subject = sd_subject
    )

    suppressWarnings(wilcox.test(dat$treated, dat$control, paired = TRUE, alternative = "two.sided")$p.value)
  })

  mean(pvals < alpha)
}
```

Build a grid over cohort size, mean shift, and subject variance:

```{r run-power-grid}
# Anchor values. You can replace these with posterior means from a specific fitted cluster.
mu0 <- 0.20
phi0 <- 30

grid <- tidyr::expand_grid(
  n_subject = seq(10, 80, by = 10),
  mean_shift = seq(0.00, 0.20, by = 0.02),
  sd_subject = c(0.1, 0.3, 0.6, 0.9)
)

set.seed(20260426)
power_grid <- grid %>%
  mutate(
    power = pmap_dbl(
      list(n_subject, mean_shift, sd_subject),
      ~ power_in_bin(
        n_subject = ..1,
        mu_control = mu0,
        mean_shift = ..2,
        phi = phi0,
        sd_subject = ..3,
        n_sims = 300
      )
    )
  )

power_grid
```

Visualize power landscapes:

```{r plot-power-grid, fig.height=6, fig.width=10}
ggplot(power_grid, aes(x = n_subject, y = mean_shift, fill = power)) +
  geom_tile() +
  facet_wrap(~ sd_subject, nrow = 1, labeller = label_both) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  labs(
    title = "Simulation-Based Power Grid (Paired Two-Sided Wilcoxon)",
    x = "Number of subjects (paired control/treated)",
    y = "Mean proportion shift (treated - control)",
    fill = "Power"
  ) +
  theme_minimal(base_size = 12)
```

## 8) Optional: Use Posterior Draws to Drive Power Simulation

A stronger workflow is to sample `(mu_control, treatment effect, phi, sd_subject)` from model posterior draws per cluster, then propagate uncertainty into power estimates.

```{r posterior-driven-power, eval=FALSE}
# Example for one fitted cluster:
cl <- names(fits)[1]
fit <- fits[[cl]]

draws <- as_draws_df(fit)

# You may need to adapt treatment coefficient naming by factor level coding.
# Here we assume b_treatmenttreated exists.

posterior_power <- replicate(200, {
  i <- sample(seq_len(nrow(draws)), 1)

  b0 <- draws$b_Intercept[i]
  bt <- draws$b_treatmenttreated[i]
  phi <- draws$phi[i]
  sd_subj <- draws$sd_subject__Intercept[i]

  mu_control <- plogis(b0)
  mu_treated <- plogis(b0 + bt)

  power_in_bin(
    n_subject = 40,
    mu_control = mu_control,
    mean_shift = mu_treated - mu_control,
    phi = phi,
    sd_subject = sd_subj,
    n_sims = 200
  )
})

summary(posterior_power)
```

## 9) Limitations, Comparison to G*Power, and Why Simulation

### Key limitations in this workflow

- Cluster proportions are compositional (sum to 1). Modeling each cluster independently with beta models ignores simplex constraints.
- Metadata quality in public integrated datasets varies (missing IDs, harmonization artifacts, uneven sampling).
- Wilcoxon tests target rank-based shifts and do not directly model compositional dependencies or hierarchical count generation.
- Power estimates depend strongly on assumptions for subject variance, dispersion, dropout, and treatment shift.

### Differences from G*Power

- G*Power is analytic and built around classical test families with standardized effect sizes.
- scRNA-seq proportion problems often violate simple parametric assumptions (bounded outcomes, zero inflation, hierarchy, compositionality).
- Simulation can incorporate realistic design details: subject pairing, unequal cells per sample, cluster prevalence, and non-normal distributions.

### Why choose simulation-based power

- Better alignment with real data-generating processes in single-cell studies.
- Directly stress-tests design sensitivity across plausible biological and technical variability.
- Produces interpretable design maps (power grids) rather than a single point estimate.
- Naturally extends to posterior predictive workflows from Bayesian models.

## 10) Next Practical Steps

1. Verify and lock exact metadata column names in `.obs` for subject/treatment/cluster.
2. Fit cluster-specific models for biologically relevant clusters.
3. Replace anchor values (`mu0`, `phi0`, variance) with posterior draws per cluster.
4. Export final power grid tables and figures for study design documentation.
